---
title: "Final Project"
author: "CQ"
date: "11/22/2019"
output: html_document
---
setwd
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
will create two new columns, 
gather()

```

#Load Packages
```{r}
# Use this to somehow put excel sheets to github to get a url to refrence
#install.packages("piggyback")
#library("piggyback")
#Need to have the github token explained again.
#pb_download("filename.rds")
#readRDS <- readRDS("filename.rds")
#save object <- saveRDS(all_results.rds) *Need to ignore in push to github.
#piggyback::pb_upload("filename.rds")

library("tidyverse")
library("tidyr")
library("readxl")
library("dplyr")
library("lubridate")
library("ggplot2")

```
setwd("/Users/chaunceyquam/Desktop/SeniorThesis/RProjectSeniorThesis/)
#Load in data from CSV files and create one table.
```{r}
df_1 <- read_csv("CSV data files/1929501.csv", col_types = cols(TMAX = col_integer(), TMIN = col_integer()),col_names=TRUE)
df_2 <- read_csv("CSV data files/1929502.csv", col_types = cols(TMAX = col_integer(), TMIN = col_integer()),col_names=TRUE)
df_3 <- read_csv("CSV data files/1929482.csv", col_types = cols(TMAX = col_integer(), TMIN = col_integer()),col_names=TRUE)
df_4 <- read_csv("CSV data files/1929489.csv", col_types = cols(TMAX = col_integer(), TMIN = col_integer()),col_names=TRUE)
df_5 <- read_csv("CSV data files/1929497.csv", col_types = cols(TMAX = col_integer(), TMIN = 
                   col_integer()),col_names=TRUE)


all_data <- rbind(df_1, df_2, df_3, df_4, df_5)
all_data
```

#Need to remove all NA variables. This omits rows with any missing values for Temperature
```{r}

all_data <- all_data[complete.cases(all_data$TMAX), ]
all_data
#Seperate the full date into three columns for year, month, day.
all_data <- separate(all_data,DATE, c("YEAR", "MONTH", "DAY"),"-")
all_data
```

#Get name of town isolated. 
#Doing this by first taking singular names column and parcing into columns.
#Then drop added columns becides CITY and STATE column
#Add first column and second together to get complete town name in CITY Column
```{r}
all_data <- separate(all_data, NAME, c("CITY","STATE","COUNTRY", "x1", "x2", "x3", "x4"), 
    sep = "([\\ \\,])")
all_data


all_data <- subset(all_data, select = -c(STATE, COUNTRY,x1,x2,x3,x4))
all_data


all_data <- unite(all_data, CITY, select =c(CITY), sep = " ")
all_data
#Check to see what every city name is.
unique(all_data$CITY) 
length(unique(all_data$CITY))
all_data 
```

#Selecting times when temp is above 105 for each city
```{r}
stat_group <- all_data %>%
  arrange(CITY) %>%
  arrange(YEAR) %>%
  filter(TMAX>105)
stat_group
unique(stat_group$CITY)

```


#Sum of times the temp was recorded to be above 105 in the selected cities in the given year.


```{r}
cities_w_CRS <- all_data %>%
  select(CITY, LATITUDE, LONGITUDE)

  
cities_w_CRS <- unique(cities_w_CRS)
cities_w_CRS
```

#ASK JAHON: LOOK INTO CREATING A VECTOR FOR THESE INSTEAD AND JOINING TOGETHER
#Creating individual tables with count of increased temp, by city, by year and joining together. 
```{r}
t_max_1989 <- all_data %>%
  filter(YEAR=="1990", TMAX>=105)
  count(CITY) %>%
  rename("2001" = n)
t_max_1989

t_max_1990 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1990",] %>%
  count(CITY) %>%
  rename("1990" = n)
t_max_1991 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1991",] %>%
  count(CITY) %>%
  rename("1991" = n)
t_max_1992 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1992",] %>%
  count(CITY) %>%
  rename("1992" = n)
t_max_1993 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1993",] %>%
  count(CITY) %>%
  rename("1993" = n)
t_max_1994 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1994",] %>%
  count(CITY) %>%
  rename("1994" = n)
t_max_1995 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1995",] %>%
  count(CITY) %>%
  rename("1995" = n)
t_max_1996 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1996",] %>%
  count(CITY) %>%
  rename("1996" = n)
t_max_1997 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1997",] %>%
  count(CITY) %>%
  rename("1997" = n)
t_max_1998<- all_data[all_data$TMAX>=105 & all_data$YEAR == "1998",] %>%
  count(CITY) %>%
  rename("1998" = n)
t_max_1999 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "1999",] %>%
  count(CITY) %>%
  rename("1999" = n)
t_max_2000 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2000",] %>%
  count(CITY) %>%
  rename("2000" = n)
t_max_2001 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2001",] %>%
  count(CITY) %>%
  rename("2000" = n)
t_max_2002 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2002",] %>%
  count(CITY) %>%
  rename(2001= n)
t_max_2003 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2003",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2004 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2004",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2005 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2005",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2006 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2006",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2007 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2007",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2008 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2008",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2009 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2009",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2010 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2010",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2011 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2011",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2012 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2012",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2013 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2013",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2014 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2014",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2015 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2015",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2016 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2016",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2017 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2017",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)
t_max_2018 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2018",] %>%
  count(CITY) %>%
  rename("Count of Days Above 105" = n)

t_max_test <- left_join(cities_w_CRS, t_max_1990, by = 'CITY') %>% 
  group_by(CITY)
t_max_test <- left_join(t_max_1989, t_max_1990, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1991, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1992, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1993, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1994, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1995, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1996, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1997, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1998, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_1999, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2000, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2001, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2002, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2003, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2004, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2005, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2006, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2007, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2008, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2009, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2010, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2011, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2012, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2013, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2014, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2015, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2016, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2017, by = 'CITY')
t_max_test <- left_join(t_max_test, t_max_2018, by = 'CITY')
t_max_test

  
test_1 <- unite(all_data, CITY, select =-c(CITY), sep = " ")
all_data <- subset(all_data, select = -c(STATE, COUNTRY,x1,x2,x3,x4))
all_data

tmax_w_1988 <- t_max_1989 %>%
  mutate(days_above_105= t_max_1998$n)

t_max_plot <- t_max_test %>% 
  ggplot(t_max_test, aes()))

ggplot(t_max_test, aes(x = t_max_test , y= CITY)) + geom_line(aes(y=no_smoothing), color='green') + geom_line(aes(y=lowess), color='blue') + xlab("Year") + ylab("Global_Temperature") +
theme_dark()
 


```

```{r}
t_max_2017 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2017",]
t_max_2017
t_max <- t_max_2017 %>%
  count(TMAX)
  
t_max_2017


```

```{r}
t_max_2018 <- all_data[all_data$TMAX>=105 & all_data$YEAR == "2018",]
t_max_2018
tally(t_max_2018)
unique(t_max_2018$CITY)
t_max_2018 %>%
  count(CITY)
```

```{r}
test_1 <- subset(all_data, select = -c(MONTH, DAY,TMIN, YEAR, TMAX))
test_1
```


```{r}
#Does the same thing as seperating date, but still contains column for full year, which is unnecessary.
#all_data <- all_data %>%
#    mutate(year = lubridate::year(DATE), 
#          month = lubridate::month(DATE), 
#          day = lubridate::day(DATE))


#Need to remove all NA variables.
  #First replace any empty values with NA
#all_data$TMAX[all_data$TMAX == " "] <- NA
#all_data$TMIN[all_data$TMIN == " "] <- NA
  #Omit any rows with missing values
#all_data[complete.cases(all_data), ]
#Another way to omit data 
#data_dropped_NA <- na.omit(all_data)
#data_dropped_NA

#Seperate values by year


#Tried to to this to get TMAX & TMIN into numeric, but first need to get rid of NA values.
#all_data <- as-factor(all_data$TMAX)
```

```{r}
#Seperating Name into Specific town names. Not working
dat2 = sapply(all_data$NAME, strsplit, ",")
dat2 = lapply(dat2, trimws)
dat2 = data.frame(matrix(unlist(dat2), ncol = 2, byrow = TRUE), stringsAsFactors = FALSE)

dat3 = sapply(dat2$X1, strsplit, " ")
dat3 = lapply(dat3, trimws)
dat3 = data.frame(matrix(unlist(dat3), ncol = 3, byrow = TRUE), stringsAsFactors = FALSE)
dat3

strsplit(as.character(all_data[c(1:5),3]),',')

```



```{r}

C
# Add a column to PV that you can visually check the regular expression matches.
PV$Match <- grepl(pattern = "(de|te)$", PV$Word)

# Subset PV data frame to show only TRUE matches
PV <- PV[PV$Match == FALSE, ]
```


```{r}
#To select values above a certain Temp Range
#myData[all_data$DATE >= "1970-01-01" & all_data$DATE <= "2016-06-27",]

```


```{r}
summary(all_data$NAME)
unique(all_data$NAME)
head(all_data$NAME)
setDT(test_df)
sapply(test_df, class)
as-factor(all_data$TMAX)
sapply(combined_df, class)
```

class(all_data)
dim(all_data)
names(all_data)
str(all_data)
hist(all_data$TMAX)

```{r}
tally(t_max_1989)
unique(t_max_1989$CITY)
tmax_count_1989 <-t_max_1989 %>%
  count(CITY)
tmax_count_1989
```


